---
title: "Spatiotemporal model with INLA"
author: Guillaume Blanchet & Steve Vissault
date: November 21th, 2019
output:
  html_document:
    highlight: tango
    toc: true
    toc_float: true
---

```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
knitr::opts_chunk$set(cache=F,fig.width=5,fig.height=5,dpi=125)
```

# Introduction

This document was written to explain how to build univariate spatiotemporal model with discrete time using INLA.

Note that INLA can be used with a range of data type because many different likelihood and link functions have been implemented in INLA. The different likelihood functions already implemented in INLA can be found by typing `INLA::inla.list.models("likelihood")` while the link functions can be found by typing `INLA::inla.list.models("link")`.

To illustrate how these models are constructed, sugar maple data from Eastern North America will be used.

# Load R package for the analysis

```{r, message = FALSE, warning=FALSE}
library(INLA)
library(raster)
library(sf)
```

# Data

```{r}
sugarRaw <- readRDS("assets/data/tree.rds")
climate <- readRDS("assets/data/climate_Present.rds")
region <- readRDS("assets/data/region.rds")
```

## Organize environmental data

```{r}
# Focus only on the data between 2010 and 2012
sugar <- sugarRaw[which(sugarRaw$yr > 2007),]

# Extract climate value for samples
climateSugar <- extract(climate, st_coordinates(sugar))
climate2007 <- scale(climateSugar[,1:2])

# Number of years
nyear <- length(unique(sugar$yr))
```

## Building a rough polygon of the study area

```{r}
# Extract extent of the sf data
extentRegion <- extent(sugar)

# Build polygon
polyRegion <- as(extentRegion, 'SpatialPolygons')  
```


## Step 1 - Building the mesh

The first step that needs to be carried out to construct a spatiotemporal model is to build a Delaunay triangulation, a mesh, that will serve as a basis to construct the spatial component of the model through an SPDE approach. 
Properly choosing the structure of the mesh is essential for the model estimation to be carried out properly. Detailed explanation on the dos and don'ts of constructing a mesh is presented in section 1.3 of the  [R-INLA tutorial on SPDE models](http://inla.r-inla-download.org/r-inla.org/tutorials/spde/html/). For a quick overview, take a look at the *mesh* practical. 
For our illustration, let's use the following mesh.

### Spatial mesh
```{r}
MeshSpace <- inla.mesh.2d(boundary = region, max.edge = 1,
                          min.angle = 20,
                          cutoff = 0.5,
                          offset = c(0.5,0.5),
                          crs = crs(region))
```

```{r, fig.width = 3, fig.height=3}
par(mar = c(1,1,1,1))

plot(MeshSpace, main = "", asp = 1)

# Number of knots in the mesh
MeshSpace$n
```

## Step 2 - Define the stochastic partial differential equation object

This is important to define the structure of the field on which the spatial contrain is imposed. In some ways, this serves as our "model variogram". Recall that in INLA, it is the Matérn model variogram that is used. In this respect, in addition of the mesh, we have to define the scaling parameter (`alpha`), the prior range (`prior.range`) and the prior variance (`prior.sigma`).

- `alpha`: This value needs to range between 0 and 2
- `prior.range`: This is a vector of length 2 specifying the range (first) and the uncertainty around this range (second). Note that the uncertainty around the range needs to be smaller that the range itself.
- `prior.sigma`: This is a vector of length 2 specifying the sill (first) and the uncertainty around this sill (second). Note that the uncertainty around the sill needs to be smaller that the range itself.

```{r}
SPDE <- inla.spde2.pcmatern(mesh=MeshSpace,
                            alpha=2,
                            prior.range=c(0.1, 0.01),
                            prior.sigma=c(1, 0.01))
```

### Step 3 - Priors and hyperparameters for temporal autocorrelation

```{r}
hSpec <- list(theta=list(prior='pccor1', param=c(0, 0.9)))
precPrior <- list(prior='pc.prec', param=c(1, 0.01)) 
```

### Step 4 - Index matrix 

Here, we construct an "index matrix" used to extract the values of the latent spatial field at the observation locations.

```{r}
Field <- inla.spde.make.index("field", n.spde=SPDE$n.spde,
                              n.group = nyear)
```

### Step 5 - $A$ matrix

We need to construct an $A$ matrix. Recall from the lecture that the $A$ matrix is sometimes referred to as the *observation matrix* and includes the parameters used to construct model. The discrete temporal groups are defined here. 

```{r}
# For estimation
A<-inla.spde.make.A(MeshSpace,
                    loc=st_coordinates(sugar),
                    group = sugar$yr - 2007)
```

**Important note** : The groups need to start at *1*. 

### Step 6 - Organise the $A$ matrix into a list

```{r}
Alist <- as.list(rep(23,4))
Alist[[1]] <- A
```

### Step 7 - Organise the effects (spatial autocorrelation structure and explanatory variables)

```{r}
effect<- list(Field,
              bio1 =  climate2007[,1],
              bio2 = climate2007[,2],
              Intercept = rep(1,nrow(climate2007)))
```

### Step 8 - Build `stack`

Here, we build a `stack` that will be passed to the `inla` function. Note that this is a special version of `stack` especially designed for the INLA package.

```{r}
Stack <- inla.stack(data=list(tree = sugar$n_occ),
                    A = Alist,
                    effects = effect,
                    tag="basis")
```

### Step 9 - Building the model (Finally !)

```{r, warning = FALSE}
form <- tree ~ 0 + bio1 + bio2 + Intercept +
        f(field, # Space
          model=SPDE, # Space
          group=field.group, # Time 
          control.group=list(model='ar1', hyper=hSpec)) # Time 

model <- inla(form,
              data = inla.stack.data(Stack),
              family="nbinomial",
              control.family =list(link="log", hyper=list(theta=precPrior)),
              control.predictor=list(A=inla.stack.A(Stack),
										                 compute=TRUE, link = 1),
              control.compute=list(waic=TRUE))
```

**Important note** : the argument `group` of the function `f()` refers to `group` defined in the object `Field`.  As such, the "name" given in this object needs to be the name of the "field".

### Study the output


```{r}
summary(model)
```

### Plot prediction maps

```{r}
# Dimention of the raster
rasterDim <- dim(climate)[1:2]

# Define basis of the map
mapBasis <- inla.mesh.projector(MeshSpace,
                               dims = rasterDim,
                               xlim = c(xmin(region), xmax(region)),
                               ylim = c(ymin(region), ymax(region)),
                               crs = crs(region))


# Calculate prediction
mapMean <- vector("list", length = nyear)
map.025 <- vector("list", length = nyear)
map.975 <- vector("list", length = nyear)

for(i in 1:nyear){
  mapMean[[i]] <- inla.mesh.project(mapBasis, 
                               model$summary.random$field$mean[Field$field.group == i])
  map.025[[i]] <- inla.mesh.project(mapBasis, 
                               model$summary.random$field$`0.025quant`[Field$field.group == i])
  map.975[[i]] <- inla.mesh.project(mapBasis, 
                               model$summary.random$field$`0.975quant`[Field$field.group == i])
}
  
# Transform map into a raster
rasterMean <- brick(climate, nl = 5)
raster.025 <- brick(climate, nl = 5)
raster.975 <- brick(climate, nl = 5)

for(i in 1:nyear){
  values(rasterMean)[,i] <- exp(as.vector(t(mapMean[[i]])))
  values(raster.025)[,i] <- exp(as.vector(t(map.025[[i]])))
  values(raster.975)[,i] <- exp(as.vector(t(map.975[[i]])))
}

# Mask the region
rasterMeanMask <- mask(rasterMean, region)
raster.025Mask <- mask(raster.025, region)
raster.975Mask <- mask(raster.975, region)

# Plot the results
for(i in 1:nyear){
par(mfrow = c(1,3))
plot(raster.025Mask[[i]], zlim = range(values(raster.025Mask[[i]]),
                              values(rasterMeanMask[[i]]),
                              values(raster.975Mask[[i]]), na.rm = TRUE),
     main = "2.5%")

plot(rasterMeanMask[[i]], zlim = range(values(raster.025Mask[[i]]),
                              values(rasterMeanMask[[i]]),
                              values(raster.975Mask[[i]]), na.rm = TRUE),
     main = "Mean")

plot(raster.975Mask[[i]], zlim = range(values(raster.025Mask[[i]]),
                              values(rasterMeanMask[[i]]),
                              values(raster.975Mask[[i]]), na.rm = TRUE),
     main = "97.5%")
}
```

