---
title: "Handling spatial data with R"
author: Guillaume Blanchet & Steve Vissault
date: November 19th, 2019
output:
  html_document:
    highlight: tango
    toc: true
    toc_float: true
---

# `r icon::fa("gamepad")` Practice 1

1. Download the [coastline shapefile](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/physical/ne_10m_coastline.zip) on your own computer.
2. Read the shapefile with the `st_read()` function.

If you want, you can try with your own data or download other shapefiles from https://www.naturalearthdata.com/

## Solution

```{r}
# if sf is not installed
# install.packages("sf")
library(sf)

# download the data 
download.file("http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/physical/ne_10m_coastline.zip", 
              destfile = 'assets/data/coastlines.zip')

# unzip the file
unzip(zipfile = "assets/data/coastlines.zip", 
      exdir = 'assets/data/ne-coastlines-10m')

# WARNING: You have to define you own path in destfile argument and exdir.

# Read the file
coast <- read_sf("assets/data/ne-coastlines-10m")
```
Make a visual check of the spatial R object

```{r}
# if mapview is not installed
# install.packages("mapview")
library(mapview)
mapview(coast)
```

### Hint: Manage your active directory

To import files on R from your computer, you have to set properly active working directory: where R is reading and writing files on your computer.
Some hints / functions to deal with that aspect.

```{r, eval = FALSE}
getwd() # Print the path of the active folder / working directory
setwd(choose.dir()) 
dir() # function which list all the files in the working directory.
```
1. `setwd()` is a function to set a new active directory
2. `choose.dir()` works exclusively on windows, This function opens a dialog box to set your custom working directory

# `r icon::fa("gamepad")` Practice 2


1. Download and read with `sf` the [canadian cities CSV file](https://simplemaps.com/static/data/country-cities/ca/ca.csv).
2. Select all cities from your province
3. Select the 10 most populated cities in your province

## Solution 

```{r}
# 1. Read file
ca_cities <- read.csv("https://simplemaps.com/static/data/country-cities/ca/ca.csv")
ca_cities <- st_as_sf(ca_cities, coords = c("lng", "lat"), crs = 4326)

# 2. Select all cities from your province
head(ca_cities)
qc_cities <- filter(ca_cities, admin == "Québec")
mapview(qc_cities)

# 3. Sort the cities dataframe by population
ca_cites <- ca_cities[order(ca_cities$population),]
# Print the 10 first cities
ca_cities[1:10,]
```

# `r icon::fa("gamepad")` Practice 3

1. Read with `sf` the [canadian cities CSV file](https://simplemaps.com/static/data/country-cities/ca/ca.csv).
2. Import with `getData` the canadian digital elevation model (elevation raster)
3. Project the elevation raster and the canadian cities in NAD83 (`raster::projectRaster()` & `sf::st_transform()`)
4. Make a buffer of 10 kms around each canadian cities (`sf::st_buffer()`)
5. Compute the elevation mean per buffer (`raster::extract()`) 

## Solution 

```{r}
# 1. Read file
ca_cities <- read.csv("https://simplemaps.com/static/data/country-cities/ca/ca.csv")
ca_cities <- st_as_sf(ca_cities, coords = c("lng", "lat"), crs = 4326)

# 2. Import the digital elevation model
altCAN <- getData(name = "alt", country = "CAN", path = "assets/data") # Coarse resolution

# 3 Print the current CRS of elevation raster 
proj4string(altCAN)
# Print the current CRS of the canadian cities
st_crs(ca_cities)

# Reproject ca_cities and altCan in 3347
# https://spatialreference.org/ref/epsg/3347/
# With sf
ca_cities <- st_transform(ca_cities, 3347)
# With raster
# This takes a while
altCAN <- projectRaster(altCAN, crs="+proj=lcc +lat_1=49 +lat_2=77 +lat_0=63.390675 +lon_0=-91.86666666666666 +x_0=6200000 +y_0=3000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs;")

# 4. Generate the buffers
cities_buf <- st_buffer(ca_cities, 10000)
mapview(cities_buf)

# 5. Crop elevation raster with buffers and compute the mean elevation
mu_elev_cities <- extract(altCAN, cities_buf, fun = mean, na.rm = TRUE)
head(mu_elev_cities)
length(mu_elev_cities)

# Visual check of the result
ca_cities$mu_elev_cities <- as.numeric(mu_elev_cities)
mapview(ca_cities, zcol = "mu_elev_cities", cex = "mu_elev_cities")
```