---
title: "Spatial variogram"
author: Guillaume Blanchet & Steve Vissault
date: November 21th, 2019
output:
  html_document:
    highlight: tango
    toc: true
    toc_float: true
---

# Prepare the R environment

We load the packages needed for this practice.

```{r}
# Spatial
library(sf)
library(sp) 
# Manipulation
library(dplyr)
# Analysis
library(gstat)
```

The first step is to load the data from the the CABIN Canadian Aquatic Biomonitoring Network in the Hamilton Harbour.
We then want to project in the NAD83 / Ontario MNR Lambert (more accurate and suitable for the study area).
If you want to read more about it: http://epsg.io/3161

# Load data

The `.rds` can be load in the R environment using the function `readRDS()`. The file contains an R object: the CABIN data for the Hamilton Harbour.

```{r}
hamilton <- readRDS("../data/post_process/hamilton_habitat.rds")
# Reproject the spatial data in NAD83 / Ontario MNR Lambert
ham_sf <- st_transform(hamilton, crs = 3161)
```

# Filter the data 

Here, we focus on the dependant variable TP and the year 2007.

```{r, fig.height=5, fig.width=12, eval=TRUE, fig.retina =2}
ham_sf_2005 <- ham_sf %>% 
        filter(Year.Année == 2005 & Variable == "TP")
ham_sf_2005_NoDup <- ham_sf_2005[!duplicated(ham_sf_2005$Site.Site),]

# Transform into an sp object needed for the variogram function
ham_SP_2005 <- as(ham_sf_2005_NoDup, "Spatial")
```

# Cloud variogram
## Build the cloud variogram

```{r}
maxDist <- max(dist(coordinates(ham_SP_2005)))
varioCloud <- variogram(Value.Valeur ~ 1,
                   data = ham_SP_2005,
                   cloud= TRUE,
                   cutoff = maxDist)
```

## Plot the cloud variogram

```{r}
par(mar = c(5,6.5,0.5,0.5))
plot(varioCloud$dist, varioCloud$gamma,
     xlab = "Distance (m)",
     ylab = "Variance",
     las = 1)
```

## Directional variogram 

To convert the variogram into a directional one, we have to set the argument `alpha`.
We pass a vector of angles through the `alpha` parameters: `c(0, 45, 90, 135)`.

```{r,echo = FALSE, fig.height=5, fig.width=12, eval=TRUE}
varioCloud <- variogram(Value.Valeur ~ 1,
                   data = ham_SP_2005,
                   cloud= TRUE,
                   cutoff = maxDist,
                   alpha = c(0, 45, 90, 135))

plot(varioCloud)
```

# Empirical variogram 

Emprical sometimes referred to sample variogram.

## Step 1 - Study spatial autocorrelation

```{r}
plot(varioCloud$dist, varioCloud$gamma,
     xlab = "Distance (m)",
     ylab = "Variance",
     las = 1)
```

### Visual autocorrelation check

```{r, fig.align="center", fig.width=12}
hscat(Value.Valeur ~ 1,
      data = ham_SP_2005,
      breaks = seq(0,maxDist, by = 500))
```

## Step 2 - Cloud variogram 

Construct a cloud variogram with a maximum distance of 1500m.

```{r, echo = FALSE, fig.align="center", fig.width=12, fig.height=5}
varioCloud <- variogram(Value.Valeur ~ 1,
                   data = ham_SP_2005,
                   cloud= TRUE,
                   cutoff = 1500)

plot(varioCloud$dist, varioCloud$gamma,
     xlab = "Distance (m)",
     ylab = "Variance",
     cex.lab = 1, las = 1)
```

## Step 3 - Check for outliers

Select point pairs by digitizing a region with the mouse (left mouse button adds a point, right mouse button ends)

```{r, eval=FALSE}
outliers <- plot(varioCloud, digitize = TRUE)
outliers
```

Ids of the row for the pair of points 

```{r, echo = FALSE}
outliers <- data.frame(head = c(4, 5), tail = c(6, 6))
outliers
```

## Step 4 - Remove outlier

```{r, eval= FALSE}
ham_SP_2005_clean <- ham_SP_2005[-6,]
```

```{r, echo= FALSE}
ham_SP_2005_clean <- ham_SP_2005
```

## Step 5 - Redraw the cloud variogram

```{r}
varioCloud <- variogram(Value.Valeur ~ 1, data = ham_SP_2005_clean,
                   cloud= TRUE,cutoff = 1500)
plot(varioCloud)
```

## Step 6 -  Define bin limits

```{r}
binLimit <- c(seq(0,1000,by = 100),
              seq(1200,1500, by = 200))
```

```{r}
# Construct variogram
varioEmpGood <- variogram(Value.Valeur ~ 1,
                          data = ham_SP_2005_clean,
                          cutoff = 1500,
                          boundaries = binLimit)
```

```{r}
plot(varioEmpGood, cex = 1, pch = 19)
```

# Residual empirical variogram

It is also possible to study the autocorrelation structure in the data **after** accounting for one or more covariates.

In the example below, we evaluate the autocorrelation structure after accounting for sampling effort.

```{r}
varioEmpResid <- variogram(Value.Valeur ~ SampleNumber.Numérod.échantillon.,
                           data = ham_SP_2005_clean,
                           cutoff = 1500,
                          boundaries = binLimit)

plot(varioEmpResid, cex = 1, pch = 19, col = "orange")
```

# Model variogram

In regard to step 1 and step 2, we choose to fit an exponential model variogram
with the restricted maximum likelihood (REML).

```{r, eval = TRUE}
modelIni <-  vgm(psill = 4e5, model = "Exp",
                 range = 800, nugget = 50000)

modelVario <- fit.variogram.reml(Value.Valeur ~ 1,
                                 data = ham_SP_2005_clean,
                                 model = modelIni)
```

## Model visualization

```{r}
plot(varioCloud, model = modelVario, lwd = 3)
```

# Exercice

1. Build a model variogram using the empirical variogram
  - This means using another model variogram fitting method
2. Build a model variogram with another variable
  - This may require using another model variogram
3. Build a model variogram on your own data
  - Think of the question you want to answer
  - The size of your data will influence the choice of model fitting procedure

# Krigging

Using the model variogram defined above we can now build kriging maps

## Step 1 -  Define a grid

```{r, eval = TRUE, echo = TRUE}
library(sf)
library(raster)
lake <- st_read("../data/raw/hamilton_harbor.gdb", layer="waterbody_2", quiet = TRUE)

# Align the projections
lake <- st_transform(lake, st_crs(ham_SP_2005_clean))

# Create the reference grid for model projection (using the lake extent)
r_ref <- raster(lake,res=100)
r_ref
```
As for inverse distance weighting, we build a grid of $100 \times 100$ m.

## Simple kriging

```{r}
# Simple kriging
skModel <- gstat(formula = Value.Valeur ~ 1, location = ham_SP_2005_clean,
            model = modelVario, beta = 2)

# Build interpolation map
skMap <- interpolate(object = r_ref, model = skModel, debug = 0)
```

### Draw the map

```{r}
library(mapview)
# Mask 
skMapMask <- mask(skMap, lake) 
mapview(skMapMask)
```

## Ordinary kriging
```{r}
# Ordinary kriging
okModel <- gstat(formula = Value.Valeur ~ 1, location = ham_SP_2005_clean,
            model = modelVario)

# Build interpolation map
okMap <- interpolate(object = r_ref, model = okModel, debug = 0)
```

### Draw the map

```{r}
library(mapview)
# Mask 
okMapMask <- mask(okMap, lake) 
mapview(okMapMask)
```

# Exercice

1. Build a kriging maps with another variable
  - This means using another model variogram
2. Perform universal kriging
3. Build a kriging maps on your own data
  - Think of the question you want to answer
  - The size of your data will influence the choice of model fitting procedure



