---
title: "Handling and visualizing spatiotemporal data"
author: Guillaume Blanchet & Steve Vissault
date: November 20th, 2019
output:
  html_document:
    highlight: tango
    toc: true
    toc_float: true
---


# Visualizing spatio-temporal data

## install & load packages 

```{r}
library(robis)
library(sf)
library(dplyr)
library(ggplot2)
library(raster)
library(sf)
library(spacetime)
library(sp)
library(mapview)

# if needed
# install.packages(c("robis","sf","dplyr","ggplot2","raster","sf","spacetime","sp"))
# install_github("andrewzm/STRbook")
```

## Gathering occurrences from OBIS

Explore [OBIS database](https://obis.org/) and get species occurrences for
Halibut (*Hippoglossus hippoglossus*).

```{r}
halibut <- robis::occurrence(scientificname = "Hippoglossus hippoglossus", 
               geometry="POLYGON((-64.84626027288095 48.91641792598304,-64.01129933538095 50.49356505646415,-68.82331105413095 50.11464395818477,-71.04254933538095 46.80876488134784,-64.84626027288095 48.91641792598304))"
          )
```

## Convert `halibut` dataframe to a sf object

```{r}
sf_halibut <- st_as_sf(halibut, 
                    coords = c("decimalLongitude","decimalLatitude"),
                    crs = 4326)
```

The [OBIS database](https://obis.org/) provide a lot of context on the records.

```{r}
names(sf_halibut)
```
We perform a visual check of the occurrences.

```{r}
mapview(sf_halibut)
```

## Visual representation per year

We can now build a representation of the occurrences per year, but firsts we have to extract the polygon for the area around the estuary.

```{r}
poly_ca <- raster::getData('GADM', country='CA', level=1)
sf_ca <- st_as_sf(poly_ca)
sf_estuary <- st_crop(sf_ca, st_as_sfc(st_bbox(sf_halibut)))
```

Following, we can visualize the data


```{r, fig.retina = 2, fig.align="center"}
ggplot() + 
     geom_sf(data = sf_estuary, fill = "grey50", col = "white") + 
     geom_sf(data = filter(sf_halibut, year > 2006), aes(col = depth)) + 
     facet_wrap(~year)
```

## Count observations per year with `rasterize`

If you have a large dataset with a high density of points, it is a good idea to
summarize the informations using a grid of abundance counts.

```{r, infos = FALSE}
# get number of years
years <- unique(sf_halibut$year)

# Create a reference grid
ref_grid <- raster(sf_halibut, res=0.05)

# Open empty stack
stack_count <- stack()

for(i in 1:length(years)){
     # Filter for one year (years[i])
     sf_halibut_yr <- filter(sf_halibut, year == years[i])
     
     # Prepare the geometry
     sp_halibut_yr <- as(st_geometry(sf_halibut_yr), "Spatial")

     # Count points per cell
     rs_count <- rasterize(sp_halibut_yr, ref_grid, fun = 'count')

     # Add the layer to the stack
     stack_count <- addLayer(stack_count,rs_count)
}

names(stack_count) <- years
```

## Compute mean values with `rasterize`

We load the air data [PM10 air quality data](https://github.com/SteveViss/PR-GSFE01/raw/master/sp_time_vario/assets/data/air.rds).

```{r}
raw_air <- readRDS("./assets/data/air.rds") 
# Transforming data to sf
sf_air <- st_as_sf(raw_air, coords=c("lng", "lat"), crs = 4326)
```

We select the variable that we want to compute the mean.

```{r}
sf_air_pm10 <- sf_air[,"pm10"]
# You can even subset for a specific year
sf_air$year <- format(sf_air$dates, "%Y")
sf_air_pm10_2003 <- subset(sf_air, year == 2003)
```
Then, we create the reference grid and perform the rasterize on the points data.

```{r}
ref_rs <- raster(sf_air, res = 0.1)
rs_mean <- rasterize(sf_air, ref_rs, fun = mean)
mapview::mapview(rs_mean[["pm10"]])
```

## Practice

1. Collect the occurences of your favorite species on OBIS.
     - `robis::occurrence(scientificname = "Hippoglossus hippoglossus")`
2. Visualize the distribution of this species across several years using `ggplot2`.
3. Produce count grids per year for this species.


# Spatial variability

**Hovmöller plot**

```{r, fig.retina = 2, fig.align = 'center'}
data("NOAA_df_1990", package = "STRbook")

Tmax <- filter(NOAA_df_1990,     # subset the data
              proc == "Tmax" &   # only max temperature
              month %in% 5:9 &   # May to September
              year == 1993)      # year of 1993

lim_lat <- range(Tmax$lat)        # latitude range

lat_axis <- seq(lim_lat[1],       # latitude axis
                lim_lat[2],
                length=25)

Tmax$t <- Tmax$julian - 728049     # create a new time variable
Tmax_grid <- Tmax
dists <- abs(outer(Tmax$lat, lat_axis, "-"))
Tmax_grid$lat <- lat_axis[apply(dists, 1, which.min)]

Tmax_lat_Hov <- group_by(Tmax_grid, lat, t) %>%
                summarise(z = mean(z))

library(STRbook)
Hovmoller_lat <- ggplot(Tmax_lat_Hov) +            # take data
        geom_tile(aes(x = lat, y = t, fill = z)) + # plot
        fill_scale(name = "degF") +     # add color scale
        scale_y_reverse() +             # rev y scale
        ylab("Day number (days)") +     # add y label
        xlab("Latitude (degrees)") +    # add x label
        theme_bw()                      # change theme

Hovmoller_lat
```

# Temporal variability

**Trends plots by stations**

```{r, fig.retina = 2, fig.align = 'center'}
data("NOAA_df_1990", package = "STRbook")

Tmax <- filter(NOAA_df_1990,     # subset the data
              proc == "Tmax" &   # only max temperature
              month %in% 5:9 &   # May to September
              year == 1993)      # year of 1993

Tmax_av <- group_by(Tmax, date) %>%
           summarise(meanTmax = mean(z))

gTmaxav <-
    ggplot() +
    geom_line(data = Tmax,aes(x = date, y = z, group = id),
              colour = "blue", alpha = 0.04) +
    geom_line(data = Tmax_av, aes(x = date, y = meanTmax)) +
    xlab("Month") + ylab("Maximum temperature (degF)") +
    theme_bw()

gTmaxav
```

# Transform your data into `spacetime` classes

To perform spatio-temporal variograms with `gstat`, we need to convert our data
into one of the following different `spacetime` classes of object. To build
these objects, the best way is to use the `stConstruct` function in the
`spacetime` package.

## Get access to the data

[Download PM10 air quality data](https://github.com/SteveViss/PR-GSFE01/raw/master/sp_time_vario/assets/data/air.rds)


## `spacetime` classes

- full grid (STF), a combination of any sp object and any xts object to represent all
possible locations on the implied space-time lattice;
- sparse grid (STS), as STF, but contains only the non-missing space-time combinations on a space-time lattice;
- irregular (STI), an irregular space-time data structure, where each point is allocated a spatial coordinate and a time stamp;
- simple trajectories (STT), a sequence of space-time points that form trajectories.


```{r}
library(spacetime)

raw_air <- readRDS("./assets/data/air.rds")
head(raw_air)

spaceTime_air <- stConstruct(raw_air, c("lng","lat"),"dates", crs = CRS("+init=epsg:4326"))
class(spaceTime_air)
```


```{r, fig.width = 10}
# subset 
raw_air_062005 <- subset(raw_air, dates > as.Date("2005-06-01") & dates < as.Date("2005-06-30"))

raw_air_062005 <- stConstruct(raw_air_062005, c("lng","lat"),"dates", crs = CRS("+init=epsg:4326"))

# Coerce the STI class to STF
STFDF_air_062005 <- as(raw_air_062005, "STFDF")
summary(STFDF_air_062005)

# Hovmöller plot
stplot(STFDF_air_062005, mode = "xt")

# Temporal-space variability
stplot(STFDF_air_062005[,,"pm10"], mode = "xy")
```

# Create metrics grid from points

```
library(sf)
raw_air <- readRDS("./assets/data/air.rds")
sf_air <- st_as_sf(raw_air, coords = c("lng","lat"), crs = 4326)
plot(sf_air)

grid <- st_make_grid(sf_air, 0.5)
mapview::mapview(grid)
inter <- st_intersects(grid, sf_air)


```


